DNAm Hierarchical Model - Fixed Effects Differential Methylation

### 2020-10-20 StanAnalysis_updated.Rmd
updated prior analysis for figures

stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
load(here(stanFile))
```

@@

fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2) # bimodal gaussian
post_prior_histogram <- function(postvals,mask,priorfun,bw,xl,titlestr) {
  # stacked histogram colored by gene-associated sites
  # and prior curve overlaid
  plot_df <- data.frame(val=postvals, gene_assoc=mask)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw) +
    stat_function(fun=function(x) bw*priorfun(x),geom="line",color="black") +
    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
  p1
}

# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.5*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.7*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "BetaT")
pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
# log sigmas
pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.52*dgamma(exp(x), 2.7693, 7.0029), 0.1, c(-5,5), "log(SigmaP)")
pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.7*dgamma(exp(x), 2.1457, 2.4819), 0.1, c(-5,5), "log(SigmaPT)")
pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6))

# arrange each parameter as a subplot # ,labels=c("M","B","E","P","PT","T","lP","lPT","lT")
p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT,hjust=0,nrow=3,ncol=3)
print(p_all)
rm(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT)
```

@@

post_histogram <- function(postvals,mask,bw,xl,titlestr) {
  # stacked histogram colored by gene-associated sites
  # and prior curve overlaid
  plot_df <- data.frame(val=postvals, gene_assoc=mask)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw) +
    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
}

# plot the two conservations score prob sums
pP1 <- post_histogram(prob1_full$median,genemask,0.05,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")"))
prob1_proxy = log(sigmaP_full$p50/sigmaT_full$p50)
pP2 <- post_histogram(prob1_proxy,genemask,0.05,c(-3,3),expression(" log( median" ~sigma[P]/~"median"~sigma[T]~")"))
# regularized version
L2norm = sqrt(sigmaP_full$p50^2 + sigmaT_full$p50^2)
prob1_reg = log(sigmaP_full$p50/sigmaT_full$p50) - 1/L2norm
pP3 <- post_histogram(prob1_reg,genemask,0.1,c(-10,5),"Regularized")
p_probs <- ggarrange(pP1,pP2,pP3,nrow=1,ncol=3)
print(p_probs)
```

@@

med_vals <- tibble(Name=FullAnnotation[,1], mu=mu_full$p50, betaT=betaT_full$p50, sigmaE=sigmaE_full$p50, sigmaP=sigmaP_full$p50, sigmaPT=sigmaPT_full$p50, sigmaT=sigmaT_full$p50, Rhat=lp_full$Rhat) %>% mutate(logP=log(sigmaP), logPT=log(sigmaPT), logT=log(sigmaT)) %>%
  mutate(L2norm = sqrt(sigmaP^2 + sigmaT^2)) %>%
  mutate(score_median=prob1_full$median, score_median2=log(sigmaP/sigmaT), score_reg=log(sigmaP/sigmaT * L2norm))
# score_prob=prob1_full$prob, score_mean=prob1_full$mean,

@@

# prep data by joining with FullAnnotation (columns: IlmnID, CHR, MAPINFO, RefGene_Name+Group, CpG_Island)
cat("Prep data with annotation, filter out sites Rhat>1.1 (and save)")
ptm<-proc.time()
prepped_data <- tibble(FullAnnotation[,c(1,2,3,5,6,8)]) %>% rename(Name = IlmnID) %>%
  left_join(by="Name", med_vals) %>% filter(Rhat<Rhat_cutoff)
write_csv(prepped_data, here("data-stan_rg3ps3", "prepped_data.csv"))
cat(proc.time()-ptm)

# separate by rows by genes
cat("Separate by gene")
ptm<-proc.time()
gened_data <- prepped_data %>%
  arrange(CHR, MAPINFO) %>%
  separate_rows(UCSC_RefGene_Name, sep=";") %>% 
  distinct()
write_csv(gened_data, here("data-stan_rg3ps3", "gened_data.csv"))
cat(proc.time()-ptm)

@@

cat(paste0("Summarize by gene, taking mean value of all sites in gene, filter out sites n<",n_cutoff))
ptm<-proc.time()
gene_summary <- gened_data %>%
  filter(!is.na(UCSC_RefGene_Name)) %>%
  gene_summarize(UCSC_RefGene_Name, c(mu:sigmaT,score_median:score_reg)) %>% filter(n>=n_cutoff)
cat(proc.time()-ptm)

@@

cat("Permutate data with bootstrapping")
ptm<-proc.time()
# 2nd function of methcon5
permed_data <- gene_summary %>%
  filter(n < quantile(n, 0.99)) %>%
  boot_naive(gened_data, sigmaP, UCSC_RefGene_Name, 1000) %>%
  boot_naive(gened_data, sigmaPT, UCSC_RefGene_Name, 1000) %>%
  boot_naive(gened_data, sigmaT, UCSC_RefGene_Name, 1000) %>%
  boot_naive(gened_data, score_median, UCSC_RefGene_Name, 1000) %>%
  boot_naive(gened_data, score_median2, UCSC_RefGene_Name, 1000) %>%
  boot_naive(gened_data, score_reg, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, sigmaP, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, sigmaPT, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, sigmaT, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, score_median, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, score_median2, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, score_reg, UCSC_RefGene_Name, 1000)
write_csv(permed_data, here("data-stan_rg3ps3", "permed_data.csv"))
cat(proc.time()-ptm)
#  boot_naive(gened_data, score_prob, UCSC_RefGene_Name, 1000) %>%
#  boot_naive(gened_data, score_mean, UCSC_RefGene_Name, 1000) %>%

# look at mean methylation conservation in promoters
cat("Calculate mean methylation conservation in promoters")
ptm<-proc.time()
promoted_data <- prepped_data %>%
  mutate(gene = str_extract(UCSC_RefGene_Name, "^[^;]+")) %>%
  filter(!is.na(gene)) %>%
  left_join(read_csv(here("data-raw","MethylationEPIC_v-1-0_B4.csv"), skip = 7) %>%  
              select(Name, Regulatory_Feature_Group), 
            by = "Name") %>%
  filter(Regulatory_Feature_Group == "Promoter_Associated") %>%
  group_by(gene) %>%
  summarise(sigmaP_pro = mean(sigmaP), sigmaPT_pro = mean(sigmaPT), sigmaT_pro = mean(sigmaT), 
            score_median_pro = mean(score_median), score_median2_pro = mean(score_median2), score_reg_pro = mean(score_reg))
write_csv(promoted_data, here("data-stan_rg3ps3", "promoted_data.csv"))
cat(proc.time()-ptm)
cat("Done with data preparation (BOTH)")
```

@@

```{r load_prep, echo=FALSE, include=TRUE, message=FALSE, eval=!params$prep}
cat("Param:prep=FALSE -> load saved data (BOTH)...")
gened_data <- read_csv(here("data-stan_rg3ps3", "gened_data.csv"), col_types = list("CHR" = col_character()))
prepped_data <- read_csv(here("data-stan_rg3ps3", "prepped_data.csv"))
permed_data <- read_csv(here("data-stan_rg3ps3", "permed_data.csv"))
promoted_data <- read_csv(here("data-stan_rg3ps3", "promoted_data.csv"))
cat("done")
```

@@

\newpage
\subsection{Tables comparing different groups of CpG sites:}
```{r Table1, echo=FALSE, warning=FALSE}
#, include=TRUE, results='asis', fig.keep='all', fig.show='asis', fig.height=3, fig.width=5
# Table 1
prepped_data %>%
  group_by(`Gene Associated` = !is.na(UCSC_RefGene_Name)) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
  mutate(`Gene Associated` = factor(`Gene Associated`, c(TRUE, FALSE), c("Yes", "No"))) %>%
  arrange(rev(rownames(.))) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))
#score_prob, score_mean,

prepped_data %>%
  group_by(`Island relation` = Relation_to_UCSC_CpG_Island) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
  dplyr::rename(ir = `Island relation`) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", "South "),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", "North "),
                        TRUE ~ ir)) %>%
  dplyr::rename(`Island relation` = ir) %>%
  arrange(rev(rownames(.))) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))

prepped_data %>%
  group_by(`5'UTR` = str_detect(UCSC_RefGene_Group, "5'UTR") & !is.na(UCSC_RefGene_Group)) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
  mutate(`5'UTR` = factor(`5'UTR`, c(TRUE, FALSE), c("Yes", "No"))) %>%
  arrange(rev(rownames(.))) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))

prepped_data %>%
  group_by(`TSS1500` = str_detect(UCSC_RefGene_Group, "TSS1500") & !is.na(UCSC_RefGene_Group)) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
  mutate(`TSS1500` = factor(`TSS1500`, c(TRUE, FALSE), c("Yes", "No"))) %>%
  arrange(rev(rownames(.))) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))

prepped_data %>%
  group_by(`TSS200` = str_detect(UCSC_RefGene_Group, "TSS200") & !is.na(UCSC_RefGene_Group)) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
  mutate(`TSS200` = factor(`TSS200`, c(TRUE, FALSE), c("Yes", "No"))) %>%
  arrange(rev(rownames(.))) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))
```

>>>>

stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
load(here(stanFile))

prepdatafolder = "data_final_rg3"
```

@@

fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2) # bimodal gaussian
post_prior_histogram <- function(postvals,mask,priorfun,bw,xl,titlestr) {
  # stacked histogram colored by gene-associated sites
  # and prior curve overlaid
  plot_df <- data.frame(val=postvals, gene_assoc=mask)
  p1 <- ggplot(plot_df, aes(x=val,fill=gene_assoc)) +
    #geom_histogram(aes(y=..count../length(mask)), binwidth=bw, position="dodge") +
    #stat_function(fun=function(x) bw*priorfun(x),geom="line",color="black") +
    geom_histogram(aes(y=..density..), binwidth=bw, position="dodge") +
    stat_function(fun=function(x) priorfun(x),geom="line",color="black") +
    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
  p1
}

# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.3*fBG(x,0.3557,-3.0675, 1.3027, 1.1290, 2.4110), 0.25, c(-5,5), "Mu")
pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.5*dcauchy(x,-0.04051, 0.5799), 0.25, c(-5,5), "BetaT")
pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 3.3*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 1.52*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.6*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 0.97*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
# log sigmas
pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.34*dgamma(exp(x), 2.7693, 7.0029), 0.25, c(-5,5), "log(SigmaP)")
pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.42*dgamma(exp(x), 2.1457, 2.4819), 0.25, c(-5,5), "log(SigmaPT)")
pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.22*dgamma(exp(x), 3.3643, 10.430), 0.25, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6))

# arrange each parameter as a subplot # ,labels=c("M","B","E","P","PT","T","lP","lPT","lT")
p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT,hjust=0,nrow=3,ncol=3)
print(p_all)
# ggsave("posteriors_geneassoc.svg",p_all)
rm(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT)
```

@@

post_histogram <- function(postvals,mask,bw,xl,titlestr) {
  # stacked histogram colored by gene-associated sites
  # and prior curve overlaid
  plot_df <- data.frame(val=postvals, gene_assoc=mask)
  p1 <- ggplot(plot_df, aes(x=val)) +
    #geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw, position="dodge") +
    geom_histogram(aes(y=..density..,fill=gene_assoc), binwidth=bw, position="dodge") +
    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Density")
  p1
}

# plot the two conservations score prob sums
pP1 <- post_histogram(CpGscore_full$median,genemask,0.05,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")")) + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
prob1_proxy = log(sigmaP_full$p50/sigmaT_full$p50)
pP2 <- post_histogram(prob1_proxy,genemask,0.05,c(-3,3),expression(" log( median" ~sigma[P]/~"median"~sigma[T]~")"))
# regularized version
L2norm = sqrt(sigmaP_full$p50^2 + sigmaT_full$p50^2)
prob1_reg = log(sigmaP_full$p50/sigmaT_full$p50) - 1/L2norm
pP3 <- post_histogram(prob1_reg,genemask,0.1,c(-10,5),"Regularized")
p_probs <- ggarrange(pP1,pP2,pP3,nrow=1,ncol=3)
print(p_probs)
```

@@

med_vals <- tibble(Name=FullAnnotation[,1], mu=mu_full$p50, betaT=betaT_full$p50, sigmaE=sigmaE_full$p50, sigmaP=sigmaP_full$p50, sigmaPT=sigmaPT_full$p50, sigmaT=sigmaT_full$p50, Rhat=lp_full$Rhat) %>% mutate(logP=log(sigmaP), logPT=log(sigmaPT), logT=log(sigmaT)) %>%
  mutate(score_median=CpGscore_full$median)

@@

# prep data by joining with FullAnnotation (columns: IlmnID, CHR, MAPINFO, RefGene_Name+Group, CpG_Island)
cat("Prep data with annotation, filter out sites Rhat>1.1 (and save)")
ptm<-proc.time()
prepped_data <- tibble(FullAnnotation[,c(1,2,3,5,6,8)]) %>% rename(Name = IlmnID) %>%
  left_join(by="Name", med_vals) %>% filter(Rhat<Rhat_cutoff)
write_csv(prepped_data, here(prepdatafolder, "prepped_data.csv"))
cat(proc.time()-ptm)

# separate by rows by genes
cat("Separate by gene")
ptm<-proc.time()
gened_data <- prepped_data %>%
  arrange(CHR, MAPINFO) %>%
  separate_rows(UCSC_RefGene_Name, sep=";") %>% 
  distinct()
write_csv(gened_data, here(prepdatafolder, "gened_data.csv"))
cat(proc.time()-ptm)

@@

cat(paste0("Summarize by gene, taking mean value of all sites in gene, filter out sites n<",n_cutoff))
ptm<-proc.time()
gene_summary <- gened_data %>%
  filter(!is.na(UCSC_RefGene_Name)) %>%
  gene_summarize(UCSC_RefGene_Name, c(mu:sigmaT,score_median)) %>% filter(n>=n_cutoff)
cat(proc.time()-ptm)

@@

cat("Permutate data with bootstrapping")
ptm<-proc.time()
# 2nd function of methcon5
permed_data <- gene_summary %>%
  filter(n < quantile(n, 0.99)) %>%
  boot_naive(gened_data, sigmaP, UCSC_RefGene_Name, 1000) %>%
  boot_naive(gened_data, sigmaPT, UCSC_RefGene_Name, 1000) %>%
  boot_naive(gened_data, sigmaT, UCSC_RefGene_Name, 1000) %>%
  boot_naive(gened_data, score_median, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, sigmaP, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, sigmaPT, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, sigmaT, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, score_median, UCSC_RefGene_Name, 1000)
write_csv(permed_data, here(prepdatafolder, "permed_data.csv"))
cat(proc.time()-ptm)

cat("Done with data preparation")
```

@@

```{r load_prep, echo=FALSE, include=TRUE, message=FALSE, eval=!params$prep}
cat("Param:prep=FALSE -> load saved data (BOTH)...")
gened_data <- read_csv(here(prepdatafolder, "gened_data.csv"), col_types = list("CHR" = col_character()))
prepped_data <- read_csv(here(prepdatafolder, "prepped_data.csv"))
permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))
promoted_data <- read_csv(here(prepdatafolder, "promoted_data.csv"))
cat("done")
```

@@

\newpage
\subsection{Tables comparing different groups of CpG sites:}
```{r Table1, echo=FALSE, warning=FALSE}
#, include=TRUE, results='asis', fig.keep='all', fig.show='asis', fig.height=3, fig.width=5
# Table 1
prepped_data %>%
  group_by(`Gene Associated` = !is.na(UCSC_RefGene_Name)) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
  mutate(`Gene Associated` = factor(`Gene Associated`, c(TRUE, FALSE), c("Yes", "No"))) %>%
  arrange(rev(rownames(.))) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
  tab_spanner(label = "Conservation Score", columns = vars(score_median))

prepped_data %>%
  group_by(`Island relation` = Relation_to_UCSC_CpG_Island) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
  dplyr::rename(ir = `Island relation`) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", "South "),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", "North "),
                        TRUE ~ ir)) %>%
  dplyr::rename(`Island relation` = ir) %>%
  arrange(rev(rownames(.))) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
  tab_spanner(label = "Conservation Score", columns = vars(score_median))

prepped_data %>%
  group_by(`5'UTR` = str_detect(UCSC_RefGene_Group, "5'UTR") & !is.na(UCSC_RefGene_Group)) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
  mutate(`5'UTR` = factor(`5'UTR`, c(TRUE, FALSE), c("Yes", "No"))) %>%
  arrange(rev(rownames(.))) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
  tab_spanner(label = "Conservation Score", columns = vars(score_median))

prepped_data %>%
  group_by(`TSS1500` = str_detect(UCSC_RefGene_Group, "TSS1500") & !is.na(UCSC_RefGene_Group)) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
  mutate(`TSS1500` = factor(`TSS1500`, c(TRUE, FALSE), c("Yes", "No"))) %>%
  arrange(rev(rownames(.))) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
  tab_spanner(label = "Conservation Score", columns = vars(score_median))

prepped_data %>%
  group_by(`TSS200` = str_detect(UCSC_RefGene_Group, "TSS200") & !is.na(UCSC_RefGene_Group)) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
  mutate(`TSS200` = factor(`TSS200`, c(TRUE, FALSE), c("Yes", "No"))) %>%
  arrange(rev(rownames(.))) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
  tab_spanner(label = "Conservation Scores", columns = vars(score_median))

# plot CpG island relation score distributions
prepped_data %>% select(Name, Relation_to_UCSC_CpG_Island, score_median) %>%
  dplyr::rename(ir = Relation_to_UCSC_CpG_Island, score=score_median) %>%
  mutate(ir = case_when(ir=="" ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", ""),
                        TRUE ~ ir)) %>%
  mutate(ir = factor(ir, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
  #ggplot() + geom_density(aes(x=score,y=..scaled..,color=ir)) + theme_classic() + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3)) + ylab("Normalized Density")
  ggplot() + geom_violin(aes(x=ir,y=score,fill=ir),draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
# ggsave("CpGscore_islands.svg")

# plot functional region score distributions
temp <- prepped_data %>% select(Name, UCSC_RefGene_Group, score_median) %>%
  separate_rows(UCSC_RefGene_Group, sep=";") %>%
  distinct() %>%
  dplyr::rename(fr = UCSC_RefGene_Group, score=score_median) %>%
  mutate(fr = case_when(fr=="" ~ "NA",
                        str_detect(fr, "5URT") ~ "5'UTR",
                        str_detect(fr, "3UTR") ~ "3'UTR",
                        TRUE ~ fr)) %>%
  distinct()
keepregions = c("TSS200","5'UTR","Body","3'UTR")
# temp %>%
#   filter(fr %in% keepregions) %>%
#   #mutate(fr = factor(fr, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
#   ggplot() + geom_density(aes(x=score,color=fr)) + theme_classic() + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3)) + ylab("Normalized Density")
temp %>%
  mutate(fr = factor(fr, levels = keepregions)) %>%
  filter(fr %in% keepregions) %>%
  ggplot() + geom_violin(aes(x=fr,y=score,fill=fr),draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
# ggsave("CpGscore_regions_violin.svg")
rm(temp)
```

### 2020-10-20 differential_methylation.Rmd
updated prior analysis for figures

# convert to M values, tInd
all_M <- logit(all_beta)
tInd <- tissueLabel=="T"

@@

nsamp = dim(all_beta)[2]
sd_all = sd(betaT_full$p50)
betaT_stats <- betaT_full %>% mutate(tstat = p50/(sem*sqrt(n_eff))) %>%
#  mutate(SSreg = sem*n_eff) %>%
#  mutate(SSres = SStot - SSreg) %>%
  mutate(pval = 2*pt(abs(tstat), nsamp-2, lower.tail=FALSE)) %>%
  mutate(qvalBH = p.adjust(pval, method="BH")) %>%
  mutate(zscore = p50/sd_all) %>%
  mutate(pz = 2*pnorm(abs(zscore),lower.tail=FALSE)) %>%
  mutate(qvalBHz = p.adjust(pz, method="BH"))

@@

\section{Compare to Irizarry et al 2009}
Irizarry use a similar differential methylation profiling approach in colorectal cancers (different methylation assay), and identify numerous cancer-specific diff. meth. regions (C-DMRs). Compare their list, which reports ranges of bases with single gene labels, to our differential methylation findings.
\newline Doi et al. 2009 (same group as Irizarry 2009), perform a similar analysis to determine C-DMRs in colorectal cancer. These C-DMRs are reported in the "DMR" column of the Illumina EPIC array manifest.
\newline Summarize % of overlap in each case, using specific bases of CpG sites or using gene names.
```{r compare_irizarry, echo=FALSE}
DMR_data = fit_stats$qvalBH <0.05
DMR_beta = betaT_stats$qvalBH < 0.05
CDMR_epic = (FullAnnotation$DMR == "CDMR") #| (FullAnnotation$DMR == "DMR")

DMR_list = cbind(DMR_data, CDMR_epic)
crosstab <- xtabs(~DMR_data + CDMR_epic, DMR_list)
crosstab

DMR_list = cbind(DMR_beta, CDMR_epic)
crosstab <- xtabs(~DMR_beta + CDMR_epic, DMR_list)
crosstab

chisq.test(crosstab)
fisher.test(crosstab)
OR = (crosstab[1,1]+crosstab[2,2])/(crosstab[1,2]+crosstab[2,1])
phi = (prod(c(crosstab[1,1],crosstab[2,2]))-prod(crosstab[1,2],crosstab[2,1]))/sqrt(prod(rowSums(crosstab))*prod(rowSums(t(crosstab))))
```

>>>>

# convert to M values, tInd
all_M <- logit(all_beta) # base e
#all_M <- log2(all_beta/(1-all_beta)) # base 2
tInd <- tissueLabel=="T"

@@

nsamp = dim(all_beta)[2]
sd_all = sd(betaT_full$p50)
betaT_stats <- betaT_full %>% mutate(tstat = p50/(sem*sqrt(n_eff))) %>%
#  mutate(SSreg = sem*n_eff) %>%
#  mutate(SSres = SStot - SSreg) %>%
  mutate(pval = 2*pt(abs(tstat), nsamp-2, lower.tail=FALSE)) %>%
  mutate(qvalBH = p.adjust(pval, method="BH")) %>%
  mutate(zscore = p50/sd_all) %>%
  mutate(pz = 2*pnorm(abs(zscore),lower.tail=FALSE)) %>%
  mutate(qvalBHz = p.adjust(pz, method="BH"))
```

\section{Compare to Irizarry et al 2009}
Irizarry use a similar differential methylation profiling approach in colorectal cancers (different methylation assay), and identify numerous cancer-specific diff. meth. regions (C-DMRs). Compare their list, which reports ranges of bases with single gene labels, to our differential methylation findings.
\newline Doi et al. 2009 (same group as Irizarry 2009), perform a similar analysis to determine C-DMRs in colorectal cancer. These C-DMRs are reported in the "DMR" column of the Illumina EPIC array manifest.
\newline Summarize % of overlap in each case, using specific bases of CpG sites or using gene names.
```{r compare_irizarry, echo=FALSE}
DMR_data = (fit_stats$qvalBH < 0.05) & (abs(fit_stats$coefficients.deltaM) > 1.4)
DMR_beta = (betaT_stats$qvalBH < 0.05) & (abs(betaT_stats$p50) > 1.4)
CDMR_epic = (FullAnnotation$DMR == "CDMR") #| (FullAnnotation$DMR == "DMR")

cor(fit_stats$tstat.ord,betaT_stats$tstat)
cor(fit_stats$coefficients.deltaM,betaT_stats$p50)

# DMR_list = cbind(DMR_data, CDMR_epic)
# crosstab <- xtabs(~DMR_data + CDMR_epic, DMR_list)
# crosstab

# DMR_list = cbind(DMR_beta, CDMR_epic)
# crosstab <- xtabs(~DMR_beta + CDMR_epic, DMR_list)
# crosstab

chisq.test(crosstab)
fisher.test(crosstab)
OR = (crosstab[1,1]+crosstab[2,2])/(crosstab[1,2]+crosstab[2,1])
phi = (prod(c(crosstab[1,1],crosstab[2,2]))-prod(crosstab[1,2],crosstab[2,1]))/sqrt(prod(rowSums(crosstab))*prod(rowSums(t(crosstab))))

# a few correlations between linear and betaT methods
cor(fit_stats$tstat.ord,betaT_stats$tstat)
cor(fit_stats$coefficients.deltaM,betaT_stats$p50)

@@
